---
title: "make-heatmaps5"
author: "Britt"
date: "2026-09-17"
output: html_document
---

let's start by getting ALL DEGs from the three treatments

```{r}
all.myc.degs <- as.data.frame(topTags(myc.qlf, n = Inf))
all.trt.degs <- as.data.frame(topTags(trt.qlf, n = Inf))
all.int.degs <- as.data.frame(topTags(interaction.qlf, n = Inf))
```

```{r}
print(all.myc.degs)
print(all.trt.degs)
print(all.int.degs)
```


```{r}
library(edgeR)
library(pheatmap)
```



```{r}

# 1. Extract row names from each DEG object
genes_trt <- rownames(top25.trt.degs)
genes_myc <- rownames(top25.myc.degs)
genes_int <- rownames(top25.int.degs)

# 2. Combine into a single vector and remove duplicate gene IDs
master_gene_list <- unique(c(genes_trt, genes_myc, genes_int))

# 3. Compute logCPM matrix and subset to master DEG list
logcpm <- cpm(dge.data, log = TRUE)
combined_heatmap_matrix <- logcpm[master_gene_list, ]

# 4. Build column annotation matching matrix column names
annotation_col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# -----------------------------------------------------------------------------
# 5. Display Separate Heatmaps Directly in RStudio Plot Window
# -----------------------------------------------------------------------------

group_list <- unique(annotation_col$Group)

for (grp in group_list) {
  
  # Identify sample columns belonging to current group
  group_samples <- rownames(annotation_col[annotation_col$Group == grp, , drop = FALSE])
  
  # Subset matrix for current group samples
  group_matrix <- combined_heatmap_matrix[, group_samples, drop = FALSE]
  
  # Filter out zero-variance rows
  row_variances <- apply(group_matrix, 1, var, na.rm = TRUE)
  group_matrix_clean <- group_matrix[row_variances > 0 & !is.na(row_variances), , drop = FALSE]
  
  if (nrow(group_matrix_clean) > 1) {
    # REMOVED filename argument to force on-screen rendering
    pheatmap(
      group_matrix_clean,
      scale = "row",
      border_color = NA,
      show_rownames = FALSE,
      show_colnames = TRUE,
      cluster_cols = TRUE,
      main = paste("Top DEGs - Group:", grp)
    )
  }
}
```


```{r}
library(edgeR)
library(pheatmap)

# 1. Extract interaction DEGs (gene names)
interaction_genes <- rownames(top25.int.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
int_heatmap_matrix <- logcpm[interaction_genes, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row_variances <- apply(int_heatmap_matrix, 1, var, na.rm = TRUE)
int_heatmap_matrix_clean <- int_heatmap_matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation_col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  int_heatmap_matrix_clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation_col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Top DEGs - Interaction Effect"
)
```

----------------------------------------------------
new plan - let's make 4 heatmaps...

1. heatmap of drought genes @ FDR 0.01
2. heatmap of all drought genes with highest mag change
3. heatmap of myc genes @ FDR 0.1
4. heatmap of all myc genes with highest mag change

so, we need two datasets...one with only drought genes and one with only myc genes

I already made them!

```{r}
all.myc.degs
all.trt.degs
```
 Now I can filter for certain FDR and logFC values...
 FDR first
 
```{r}
sig.trt.degs <- topTags(trt.qlf, n = Inf, p.value = 0.01)$table
sig.myc.degs <- topTags(myc.qlf, n = Inf, p.value = 0.1)$table

sig.trt.degs
sig.myc.degs
```
 and now logFC...
 
```{r}
library(dplyr)

# Filter strictly for |log2FC| >= 1
logFC.myc.degs <- all.myc.degs %>%
  filter(abs(logFC) >= 1)

logFC.trt.degs <- all.trt.degs %>%
  filter(abs(logFC) >= 1)

logFC.myc.degs
logFC.trt.degs
```
 
here are my four datasets:

```{r}
sig.myc.degs #all sig myc genes > 0.1 FDR
sig.trt.degs #all sig trt genes > 0.01 FDR
logFC.myc.degs #all genes > abs Log2FC 1
logFC.trt.degs #all genes > abs Log2FC 1
```

let's try to make heatmaps.

```{r}
library(edgeR)
library(pheatmap)

# 1. Extract DEGs (gene names)
sig.myc.genes.names <- rownames(sig.myc.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
sig.myc.heatmap.matrix <- logcpm[sig.myc.genes.names, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row.variances <- apply(sig.myc.heatmap.matrix, 1, var, na.rm = TRUE)
sig.myc.clean <- sig.myc.heatmap.matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation.col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  sig.myc.clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation.col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Myc DEGs > 0.1 FDR",
   filename = "results/new-results/sig.myc.heatmap.pdf",  # Saved to new subfolder
  width = 8,
  height = 10
)
```

```{r}
# 1. Extract DEGs (gene names)
sig.trt.genes.names <- rownames(sig.trt.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
sig.trt.heatmap.matrix <- logcpm[sig.trt.genes.names, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row.variances <- apply(sig.trt.heatmap.matrix, 1, var, na.rm = TRUE)
sig.trt.clean <- sig.trt.heatmap.matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation.col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  sig.trt.clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation.col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Trt DEGs > 0.01 FDR",
   filename = "results/new-results/sig.trt.heatmap.pdf",  # Saved to new subfolder
  width = 8,
  height = 10
)
```

```{r}
# 1. Extract DEGs (gene names)
logFC.myc.genes.names <- rownames(logFC.myc.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
logFC.myc.heatmap.matrix <- logcpm[logFC.myc.genes.names, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row.variances <- apply(logFC.myc.heatmap.matrix, 1, var, na.rm = TRUE)
logFC.myc.clean <- logFC.myc.heatmap.matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation.col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  logFC.myc.clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation.col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Myc DEGs > 1 logFC",
   filename = "results/new-results/logFC.myc.heatmap.pdf",  # Saved to new subfolder
  width = 8,
  height = 10
)
```


```{r}
# 1. Extract DEGs (gene names)
logFC.trt.genes.names <- rownames(logFC.trt.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
logFC.trt.heatmap.matrix <- logcpm[logFC.trt.genes.names, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row.variances <- apply(logFC.trt.heatmap.matrix, 1, var, na.rm = TRUE)
logFC.trt.clean <- logFC.trt.heatmap.matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation.col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  logFC.trt.clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation.col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Trt DEGs > 1 logFC",
)

# 1. Create the nested results/new-results folder
dir.create("results/new-results", recursive = TRUE, showWarnings = FALSE)

# 2. Reset any active graphic connections from previous attempts
while (!is.null(dev.list())) { dev.off() }

```


```{r}
# 3. Export heatmap to the new subfolder
pheatmap(
  logFC.trt.clean,
  scale = "row",
  annotation_col = annotation.col,
  border_color = NA,
  show_rownames = FALSE,
  show_colnames = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Trt DEGs > 1 logFC",
  filename = "results/new-results/logFC.trt.heatmap.pdf",  # Saved to new subfolder
  width = 8,
  height = 10
)
```



let's save heatmaps

```{r}
# Save as high-resolution PNG
pheatmap(
  mat_scaled,
  filename = "results/heatmap.png",
  width = 8,
  height = 10,
  res = 300  # Sets resolution to 300 DPI
)
```




