---
title: "Final DFD Analysis"
author: "Britt"
date: "2026-04-07"
output: html_document
---

```{r}
PM.counts <- read.table("Psme_RNASeq_July2024_Psme.1_0_counts.txt", header = TRUE) 
PM.counts
```

```{r}
PM.counts <- PM.counts[rowSums(PM.counts[,-1] > 10) >= 3,]
PM.counts
```

```{r}
library(tibble)
PM.counts.noid <- as_tibble(PM.counts)
PM.counts.noid
```

```{r}
library(tidyverse)
sample.description <- tibble(sample=colnames(TC.counts.noid))
sample.description <- sample.description %>%
  mutate(
    myc = c("NM", "NM", "NM", "NM", "M", "NM", "M", "M", "M", "NM", "M", "NM", "NM", "M", "NM", "M", "NM", "NM", "NM", "M", "NA"),  
    trt = c("C", "C", "T", "C", "C", "T", "T", "T", "C", "C", "C", "C", "T", "T", "C", "C", "C", "T", "T", "T", "NA"),  
    group = paste(myc, trt, sep = "_")
  )
sample.description <- sample.description %>%
  slice(1:(n() - 1))
sample.description <- sample.description %>%
  mutate(myc=factor(myc, levels = c("NM","M")), 
         trt=factor(trt,levels = c("C","T"))) # setting the levels in this way makes "C and NM" the reference
sample.description
```


```{r}
library(edgeR)

counts.matrix.PM <- PM.counts %>% as.matrix()

#remove unassigned row
counts.matrix.PM <- counts.matrix.PM[, -21]

counts.matrix.PM
```

```{r}
dge.data <- DGEList(counts=counts.matrix.PM, 
                    group=sample.description$group)
dim(dge.data) 
dge.data <- calcNormFactors(dge.data, method = "TMM")
dge.data$samples # look at the normalization factors
```

(make ordination here)

```{r}
# Ensure levels are set correctly
sample.description$myc <- relevel(factor(sample.description$myc), ref = "NM")
sample.description$trt <- relevel(factor(sample.description$trt), ref = "C")
```


```{r}
#sample.description is the original descriptive dataset (describes our samples but doesn't have counts data)

interaction.design <- model.matrix(~myc*trt,data = sample.description)
rownames(interaction.design) <- sample.description$sample
interaction.design
```
```{r}
# Modern, clean workflow for the Interaction model
dge.data <- estimateDisp(dge.data, interaction.design)

# View the dispersion estimates
summary(dge.data$tagwise.dispersion)
```



```{r}
# 1. Fit using the QL model instead of glmFit
fit <- glmQLFit(dge.data, interaction.design)

# 2. Test using the QL F-test instead of glmLRT
# Treatment Effect
trt.qlf <- glmQLFTest(fit, coef = "trtT")

# Mycorrhization Effect (spM)
myc.qlf <- glmQLFTest(fit, coef = "mycM")

# Interaction Effect (The Synergy)
interaction.qlf <- glmQLFTest(fit, coef = "mycM:trtT")
```


```{r}
summary(decideTests(myc.qlf, p.value=0.06))
summary(decideTests(trt.qlf, p.value=0.01))
summary(decideTests(interaction.qlf, p.value=0.5))
```

```{r}
top10.trt <- topTags(trt.qlf)
top10.myc <- topTags(myc.qlf) 
top10.interaction <- topTags(interaction.qlf)

top10.trt
top10.myc
top10.interaction
```
Notes:

interaction: no sig genes, top genes at 0.5 FDR and higher (not 0.25 threshold Laura and I discussed)
myc: sig genes at above 0.05 FDR, set at 0.06 produces sig DEGs
trt: sig genes at 0.01 FDR


```{r}
# Visualizing the dispersion/BCV
plotBCV(dge.data)
```

```{r}
# This will show you if you have an outlier sample
plotMDS(dge.data, col=as.numeric(sample.description$myc))
```

```{r}
mdsvals <- plotMDS(dge.data, plot = FALSE) #using this one for paper

var1 <- round(mdsvals$var.explained[1] * 100, 1)
var2 <- round(mdsvals$var.explained[2] * 100, 1)

tibble(
  x = mdsvals$x,
  y = mdsvals$y,
  sample = rownames(dge.data$samples)
) %>%
  inner_join(sample.description, by = "sample") %>%
  ggplot(aes(x = x, y = y, color = trt, shape = myc)) +
  geom_point(size = 3) +
  labs(
    x = paste0("Dimension 1 (", var1, "%)"),
    y = paste0("Dimension 2 (", var2, "%)"),
    color = "Drought",
    shape = "Mycorrhization"
  ) +
  theme_classic()
```

Notes: PCA does not change with new analysis. Still not looking great.

------------------------------------------------
9/7/26 let's make heatmaps

let's start by getting ALL DEGs from the three treatments

```{r}
all.myc.degs <- as.data.frame(topTags(myc.qlf, n = Inf))
all.trt.degs <- as.data.frame(topTags(trt.qlf, n = Inf))
all.int.degs <- as.data.frame(topTags(interaction.qlf, n = Inf))
```

```{r}
print(all.myc.degs)
print(all.trt.degs)
print(all.int.degs)
```


```{r}
library(edgeR)
library(pheatmap)
```

```{r}
# Filter for significant DEGs (FDR < 0.05) and select top 25 by absolute logFC
sig.trt.degs <- subset(all.trt.degs, FDR < 0.01)
top25.trt.degs <- sig.trt.degs[order(abs(sig.trt.degs$logFC),decreasing = TRUE), ][1:25, ]

sig.myc.degs <- subset(all.myc.degs, FDR < 0.06)
top25.myc.degs <- sig.myc.degs[order(abs(sig.myc.degs$logFC),decreasing = TRUE), ][1:25, ]

top25.int.degs <- all.int.degs[order(abs(all.int.degs$logFC),decreasing = TRUE), ][1:25, ]
```

```{r}
top25.int.degs
top25.myc.degs
top25.trt.degs
```

```{r}
# 1. Extract the gene names (row names) from each of the three top 25 objects
genes_trt <- rownames(top25.trt.degs)
genes_myc <- rownames(top25.myc.degs)
genes_int <- rownames(top25.int.degs)
```


```{r}
# 2. Combine them into a single vector and remove duplicates using unique()
master_gene_list <- unique(c(genes_trt, genes_myc, genes_int))
```


```{r}
logcpm <- cpm(dge.data, log = TRUE)
# 3. Use the single vector to subset your logCPM expression matrix for pheatmap
combined_heatmap_matrix <- logcpm[master_gene_list, ]
```

```{r}
# Option B: If sample description is already inside edgeR's dge_object
annotation_col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm) # Must match the matrix column names
)
```



```{r}

# 4. Draw the combined heatmap
pheatmap(
  combined_heatmap_matrix,
  scale = "row",
  annotation_col = annotation_col,
  show_rownames = TRUE,
  main = "Combined Top DEGs Across All Comparisons"
)
```

```{r}
pheatmap(
  combined_heatmap_matrix,
  scale = "row",                     # Essential: Centers/scales row expression (Z-score)
  annotation_col = annotation_col,  # Keeps sample group annotations
  show_rownames = FALSE,             # Hides compressed gene IDs (or set TRUE and export as taller PDF)
  show_colnames = TRUE,              # Keeps sample IDs
  border_color = NA,                 # Removes grey cell borders for a cleaner look
  fontsize_col = 10,                 # Adjusts column label font size
  main = "Combined Top DEGs Across All Comparisons"
)
```


```{r}
library(edgeR)
library(pheatmap)

# 1. Extract row names from each DEG object
genes_trt <- rownames(top25.trt.degs)
genes_myc <- rownames(top25.myc.degs)
genes_int <- rownames(top25.int.degs)

# 2. Combine into a single vector and remove duplicate gene IDs
master_gene_list <- unique(c(genes_trt, genes_myc, genes_int))

# 3. Compute logCPM matrix and subset to master DEG list
logcpm <- cpm(dge.data, log = TRUE)
combined_heatmap_matrix <- logcpm[master_gene_list, ]

# 4. Build column annotation matching matrix column names
annotation_col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# -----------------------------------------------------------------------------
# 5. Display Separate Heatmaps Directly in RStudio Plot Window
# -----------------------------------------------------------------------------

group_list <- unique(annotation_col$Group)

for (grp in group_list) {
  
  # Identify sample columns belonging to current group
  group_samples <- rownames(annotation_col[annotation_col$Group == grp, , drop = FALSE])
  
  # Subset matrix for current group samples
  group_matrix <- combined_heatmap_matrix[, group_samples, drop = FALSE]
  
  # Filter out zero-variance rows
  row_variances <- apply(group_matrix, 1, var, na.rm = TRUE)
  group_matrix_clean <- group_matrix[row_variances > 0 & !is.na(row_variances), , drop = FALSE]
  
  if (nrow(group_matrix_clean) > 1) {
    # REMOVED filename argument to force on-screen rendering
    pheatmap(
      group_matrix_clean,
      scale = "row",
      border_color = NA,
      show_rownames = FALSE,
      show_colnames = TRUE,
      cluster_cols = TRUE,
      main = paste("Top DEGs - Group:", grp)
    )
  }
}
```


```{r}
library(edgeR)
library(pheatmap)

# 1. Extract interaction DEGs (gene names)
interaction_genes <- rownames(top25.int.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
int_heatmap_matrix <- logcpm[interaction_genes, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row_variances <- apply(int_heatmap_matrix, 1, var, na.rm = TRUE)
int_heatmap_matrix_clean <- int_heatmap_matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation_col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  int_heatmap_matrix_clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation_col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Top DEGs - Interaction Effect"
)
```

----------------------------------------------------
new plan - let's make 4 heatmaps...

1. heatmap of drought genes @ FDR 0.01
2. heatmap of all drought genes with highest mag change
3. heatmap of myc genes @ FDR 0.1
4. heatmap of all myc genes with highest mag change

so, we need two datasets...one with only drought genes and one with only myc genes

I already made them!

```{r}
all.myc.degs
all.trt.degs
```
 Now I can filter for certain FDR and logFC values...
 FDR first
 
```{r}
sig.trt.degs <- topTags(trt.qlf, n = Inf, p.value = 0.01)$table
sig.myc.degs <- topTags(myc.qlf, n = Inf, p.value = 0.1)$table

sig.trt.degs
sig.myc.degs
```
 and now logFC...
 
```{r}
library(dplyr)

# Filter strictly for |log2FC| >= 1
logFC.myc.degs <- all.myc.degs %>%
  filter(abs(logFC) >= 1)

logFC.trt.degs <- all.trt.degs %>%
  filter(abs(logFC) >= 1)

logFC.myc.degs
logFC.trt.degs
```
 
here are my four datasets:

```{r}
sig.myc.degs #all sig myc genes > 0.1 FDR
sig.trt.degs #all sig trt genes > 0.01 FDR
logFC.myc.degs #all genes > abs Log2FC 1
logFC.trt.degs #all genes > abs Log2FC 1
```

let's try to make heatmaps.

```{r}
library(edgeR)
library(pheatmap)

# 1. Extract DEGs (gene names)
sig.myc.genes.names <- rownames(sig.myc.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
sig.myc.heatmap.matrix <- logcpm[sig.myc.genes.names, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row.variances <- apply(sig.myc.heatmap.matrix, 1, var, na.rm = TRUE)
sig.myc.clean <- sig.myc.heatmap.matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation.col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  sig.myc.clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation.col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Myc DEGs > 0.1 FDR",
   filename = "results/new-results/sig.myc.heatmap.pdf",  # Saved to new subfolder
  width = 8,
  height = 10
)
```

```{r}
# 1. Extract DEGs (gene names)
sig.trt.genes.names <- rownames(sig.trt.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
sig.trt.heatmap.matrix <- logcpm[sig.trt.genes.names, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row.variances <- apply(sig.trt.heatmap.matrix, 1, var, na.rm = TRUE)
sig.trt.clean <- sig.trt.heatmap.matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation.col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  sig.trt.clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation.col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Trt DEGs > 0.01 FDR",
   filename = "results/new-results/sig.trt.heatmap.pdf",  # Saved to new subfolder
  width = 8,
  height = 10
)
```

```{r}
# 1. Extract DEGs (gene names)
logFC.myc.genes.names <- rownames(logFC.myc.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
logFC.myc.heatmap.matrix <- logcpm[logFC.myc.genes.names, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row.variances <- apply(logFC.myc.heatmap.matrix, 1, var, na.rm = TRUE)
logFC.myc.clean <- logFC.myc.heatmap.matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation.col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  logFC.myc.clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation.col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Myc DEGs > 1 logFC",
   filename = "results/new-results/logFC.myc.heatmap.pdf",  # Saved to new subfolder
  width = 8,
  height = 10
)
```


```{r}
# 1. Extract DEGs (gene names)
logFC.trt.genes.names <- rownames(logFC.trt.degs)

# 2. Compute logCPM and subset matrix for interaction genes only
logcpm <- cpm(dge.data, log = TRUE)
logFC.trt.heatmap.matrix <- logcpm[logFC.trt.genes.names, ]

# 3. Filter out zero-variance rows (prevents scaling errors)
row.variances <- apply(logFC.trt.heatmap.matrix, 1, var, na.rm = TRUE)
logFC.trt.clean <- logFC.trt.heatmap.matrix[row_variances > 0 & !is.na(row_variances), ]

# 4. Build group annotations
annotation.col <- data.frame(
  Group = dge.data$samples$group,
  row.names = colnames(logcpm)
)

# 5. Render the interaction heatmap
pheatmap(
  logFC.trt.clean,
  scale = "row",                    # Standardize row expression (Z-score)
  annotation_col = annotation.col,  # Show sample groups at top
  border_color = NA,               # Clean tile display
  show_rownames = FALSE,           # Hide dense row labels
  show_colnames = TRUE,            # Show sample names
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Trt DEGs > 1 logFC",
)

# 1. Create the nested results/new-results folder
dir.create("results/new-results", recursive = TRUE, showWarnings = FALSE)

# 2. Reset any active graphic connections from previous attempts
while (!is.null(dev.list())) { dev.off() }

```


```{r}
# 3. Export heatmap to the new subfolder
pheatmap(
  logFC.trt.clean,
  scale = "row",
  annotation_col = annotation.col,
  border_color = NA,
  show_rownames = FALSE,
  show_colnames = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Trt DEGs > 1 logFC",
  filename = "results/new-results/logFC.trt.heatmap.pdf",  # Saved to new subfolder
  width = 8,
  height = 10
)
```



let's save heatmaps

```{r}
# Save as high-resolution PNG
pheatmap(
  mat_scaled,
  filename = "results/heatmap.png",
  width = 8,
  height = 10,
  res = 300  # Sets resolution to 300 DPI
)
```

------------------------------------------------------------

next: 
fix heatmap5 file (make it coherent)
make interpro script





