---
title: "Bradley DEGs"
author: "Britt"
date: "2024-09-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Download counts tables from Bogar server:

```{bash, eval = FALSE}
#in personal computer terminal:

scp -P 22345 bklong@bogarprecision.plb.ucdavis.edu:~/DemRNAanalysis/BioShare/04-Counts/Psme_RNASeq_July2024_Psme.1_0_counts.txt /c/Users/britt/Downloads

scp -P 22345 bklong@bogarprecision.plb.ucdavis.edu:~/DemRNAanalysis/BioShare/04-Counts/Psme_RNASeq_July2024_Trucit1_1_counts.txt /c/Users/britt/Downloads
```

The counts tables are now on my personal computer where R can access them.

### Starting Analysis for Both:

**Basic steps for DGE Analysis**

1.  Load the RNAseq counts.
2.  Normalize the counts.
3.  QC the counts.
4.  Create a data frame that describes the experiment.
5.  Determine the dispersion parameter (how over-dispersed is the data?)
6.  Fit a statistical model for gene expression as a function of experimental parameters.
7.  Test the significance of experimental parameters for explaining gene expression.
8.  Examine result.

Let's take a look at the data:

```{r}
PM_counts <- read.table("Psme_RNASeq_July2024_Psme.1_0_counts.txt", header = TRUE) 

PM_counts
```

```{r}
TC_counts <- read.table("Psme_RNASeq_July2024_Trucit1_1_counts.txt", header = TRUE) 
TC_counts
```

Filter the counts data so that you only retain genes that were expressed:

```{r}
#retain only the reads that were over 10 count in at least three samples

TC_counts <- TC_counts[rowSums(TC_counts[,-1] > 10) >= 3,]

TC_counts

PM_counts <- PM_counts[rowSums(PM_counts[,-1] > 10) >= 3,]

PM_counts
```

Remove Transcript ID column (convert dataframe to tibble).

```{r}
library(tibble)

TC_counts_noid <- as_tibble(TC_counts)

TC_counts_noid

PM_counts_noid <- as_tibble(PM_counts)

PM_counts_noid
```

Log transform the counts data.

```{r}
transformed.TC.counts <- log2(TC_counts_noid + 1) 
transformed.PM.counts <- log2(PM_counts_noid + 1)
```

Correlation table and pairs:

```{r}
cor.table.TC <- cor(TC_counts_noid)
cor.table.PM <- cor(PM_counts_noid)
```

```{r}
pairs(transformed.PM.counts)
pairs(transformed.TC.counts)
```

```{r}
library(tidyverse)

cor.table.PM %>% gplots::heatmap.2(dendrogram="row", trace = "none", col=viridis::viridis(25, begin=.25), margins=c(4,4))
```

```{r}
cor.table.TC %>% gplots::heatmap.2(dendrogram="row", trace = "none", col=viridis::viridis(25, begin=.25), margins=c(4,4))
```

```{r}
#make tibble with descriptions from counts data table
sample.description <- tibble(sample=colnames(TC_counts_noid))

sample.description
```

```{r}
sample.description <- sample.description %>%
  mutate(
    sp = c("NM", "NM", "NM", "NM", "ST", "NM", "TC", "ST", "ST", "NM", "TC", "NM", "NM", "ST", "NM", "ST", "NM", "NM", "NM", "ST", "NA"),  
    trt = c("C", "C", "T", "C", "C", "T", "T", "T", "C", "C", "C", "C", "T", "T", "C", "C", "C", "T", "T", "T", "NA"),  
    group = paste(sp, trt, sep = "_")
  )

sample.description
```

```{r}
sample.description <- sample.description %>%
  slice(1:(n() - 1))
sample.description
```

```{r}
sample.description <- sample.description %>%
  mutate(sp=factor(sp), 
         trt=factor(trt,levels = c("C","T"))) # setting the levels in this way makes "C" the reference  

sample.description
```

### Tree Counts Analysis:

```{r}
library(edgeR)

counts.matrix.PM <- PM_counts %>% as.matrix()

counts.matrix.PM
```

```{r}
#remove unassigned row

counts.matrix.PM <- counts.matrix.PM[, -21]

counts.matrix.PM
```

```{r}
dge.data.PM <- DGEList(counts=counts.matrix.PM, 
                    group=sample.description$group)
dim(dge.data.PM) 
dge.data.PM <- calcNormFactors(dge.data.PM, method = "TMM")
dge.data.PM$samples # look at the normalization factors
```

MDS plot:

```{r}
plotMDS(dge.data.PM)
```

Nicer MDS:

```{r}
mdsvals <- plotMDS(dge.data.PM, plot = FALSE) # get the MDS values for plotting

tibble(x=mdsvals$x, y=mdsvals$y, sample=rownames(dge.data.PM$samples)) %>%
  inner_join(sample.description) %>%
  ggplot(aes(x=x, y=y, color=trt, shape=sp)) +
  geom_point(size=3) 
```

```{r}
#does this even get used somewhere? this may be an area to ask for help in...may be a mistake

PM.counts.normal <- cpm(dge.data.PM) 

# or log2 transformed:
PM.counts.normal.log <- cpm(dge.data.PM,log = TRUE)
```

```{r}
#sample.description is the original descriptive dataset (describes our samples but doesn't have counts data)

design.PM <- model.matrix(~sp+trt,data = sample.description)
rownames(design.PM) <- sample.description$sample
design.PM
```

```{r}
#First the overall dispersion
dge.data.PM <- estimateGLMCommonDisp(dge.data.PM,design.PM,verbose = TRUE)

#Then a trended dispersion based on count level
dge.data.PM <- estimateGLMTrendedDisp(dge.data.PM,design.PM)

#And lastly we calculate the gene-wise dispersion, using the prior estimates to "squeeze" the dispersion towards the common dispersion.
dge.data.PM <- estimateGLMTagwiseDisp(dge.data.PM,design.PM)

#We can examine this with a plot
plotBCV(dge.data.PM)
```

```{r}
#this is the full model with all coefficients
#dge.data is the normalized count data
fit.PM <- glmFit(dge.data.PM, design.PM)
```

```{r}
#this model looks at genes differentially expressed between two treatments
trt.lrt.PM <- glmLRT(fit.PM,coef = "trtT")
```

```{r}
#this model between species
#there are three species...how to fix?
# Create a contrast vector
# The vector should be 1 for the intercept (reference level) and -0.5 for each non-reference level in the comparison
contrast <- c(1, -0.5, -0.5, 0)  
# Adjust according to your design matrix columns

# Perform the likelihood ratio test with the custom contrast
sp.lrt.PM <- glmLRT(fit.PM,contrast = contrast)
```

```{r}
sp.lrt.PM.TC <- glmLRT(fit.PM,coef = "spTC")
sp.lrt.PM.ST <- glmLRT(fit.PM, coef = "spST")
```

```{r}
top10.sp.PM.ST <- topTags(sp.lrt.PM.ST) # the top 10 most differentially expressed genes between the two treatments (using the above model)
top10.sp.PM.ST
```

```{r}
top10.sp.PM.TC <- topTags(sp.lrt.PM.TC) # the top 10 most differentially expressed genes between the two treatments (using the above model)
top10.sp.PM.TC
```

```{r}
top10.sp.PM <- topTags(sp.lrt.PM) # the top 10 most differentially expressed genes between the two treatments (using the above model)
top10.sp.PM
```

```{r}
top10.trt.PM <- topTags(trt.lrt.PM) # the top 10 most differentially expressed genes between the two treatments (using the above model)
top10.trt.PM
```

```{r}
summary(decideTestsDGE(trt.lrt.PM,p.value=0.01)) #This uses the FDR.  0.05 would be OK also.
```

```{r}
summary(decideTestsDGE(sp.lrt.PM,p.value=0.01)) #This uses the FDR.  0.05 would be OK also.
```

```{r}
summary(decideTestsDGE(sp.lrt.PM.TC,p.value=0.01)) #This uses the FDR.  0.05 would be OK also.
```

```{r}
summary(decideTestsDGE(sp.lrt.PM.ST,p.value=0.01)) #This uses the FDR.  0.05 would be OK also.
```

heatmap attempt:

```{r}
library(pheatmap)
pheatmap(PM.counts.normal.log, scale = "row")
```

make new dataframe to make heatmap that has treatments included in sample labels:

```{r}
PM.counts.normal.log.named <- PM.counts.normal.log
colnames(PM.counts.normal.log.named) <- c()
```

```{r}
# Subset for top genes
top_genes_trt <- rownames(topTags(trt.lrt.PM, n = 50)$table)
top50PMtrt <- PM.counts.normal.log[top_genes, ]

#write .csv file
write.csv(top50PM, file = "top50PMtrt.csv")
```

interaction DEGs for PM:

```{r}
design.PM.interaction <- model.matrix(~sp*trt,data = sample.description)

rownames(design.PM.interaction) <- sample.description$sample

design.PM.interaction
```

```{r}
#First the overall dispersion
dge.data.PM.interaction <- estimateGLMCommonDisp(dge.data.PM,design.PM.interaction,verbose = TRUE)

#Then a trended dispersion based on count level
dge.data.PM.interaction <- estimateGLMTrendedDisp(dge.data.PM.interaction,design.PM.interaction)

#And lastly we calculate the gene-wise dispersion, using the prior estimates to "squeeze" the dispersion towards the common dispersion.
dge.data.PM.interaction <- estimateGLMTagwiseDisp(dge.data.PM.interaction,design.PM.interaction)

#We can examine this with a plot
plotBCV(dge.data.PM.interaction)
```

```{r}
fit.interaction.PM <- glmFit(dge.data.PM.interaction, design.PM.interaction)
```

```{r, eval = FALSE}
#I can't figure this out

# Original interaction term name
STinteraction <- "spST:trtT"
TCinteraction <- "spTC:trtT"

# Convert it to a valid name
STinteraction <- make.names(STinteraction)
TCinteraction <- make.names(TCinteraction)

# Use this valid name in makeContrasts
contrast.matrix <- makeContrasts(
    interaction1 = STinteraction,
    interaction2 = TCinteraction,
    levels = design.PM.interaction
)

lrt.interaction.PM <- glmLRT(fit.PM, contrast = contrast.matrix)
```

```{r}
lrt.interaction.PM <- glmLRT(fit.interaction.PM, coef = "spST:trtT")
```

```{r}
lrt.interaction.PM2 <- glmLRT(fit.interaction.PM, coef = "spTC:trtT")
```

```{r}
topTags(lrt.interaction.PM) # the top 10 most differentially expressed genes
```

```{r}
topTags(lrt.interaction.PM2)
```

```{r}
summary(decideTestsDGE(lrt.interaction.PM2, p.value=0.01))
```

```{r}
summary(decideTestsDGE(lrt.interaction.PM,p.value=0.01)) #This uses the FDR.  0.05 would be OK also.
```

```{r}
#Extract genes with a FDR < 0.01 (could also use 0.05)
PM.ST.interaction <- topTags(lrt.interaction.PM,n = Inf,p.value = 0.01)$table

PM.TC.interaction <- topTags(lrt.interaction.PM2, n = Inf, p.value = 0.01)$table

#save to a file
write.csv(PM.ST.interaction,"PM.ST.interaction.csv")
write.csv(PM.TC.interaction,"PM.TC.interaction.csv")
```

```{bash, eval = FALSE}
zgrep ">PSME_30375" Psme.1_0.fa.gz
```

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\--

PM to do:

visualization for PM DEGs

annotation for PM DEGs

interaction DEGs for PM (between 3 sp and 2 trt, tricky)

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\--

### Fungal Counts Analysis:

**Basic steps for DGE Analysis**

1.  Load the RNAseq counts.
2.  Normalize the counts.
3.  QC the counts.
4.  Create a data frame that describes the experiment.
5.  Determine the dispersion parameter (how over-dispersed is the data?)
6.  Fit a statistical model for gene expression as a function of experimental parameters.
7.  Test the significance of experimental parameters for explaining gene expression.
8.  Examine result.

```{r}
counts.matrix.TC <- TC_counts %>% as.matrix()

counts.matrix.TC
```

```{r}
#remove unassigned row

counts.matrix.TC <- counts.matrix.TC[, -21]

counts.matrix.TC
```

```{r}
dge.data.TC <- DGEList(counts=counts.matrix.TC, 
                    group=sample.description$group)
dim(dge.data.TC) 
dge.data.TC <- calcNormFactors(dge.data.TC, method = "TMM")
dge.data.TC$samples # look at the normalization factors
```

MDS:

```{r}
mdsvals <- plotMDS(dge.data.TC, plot = FALSE) # get the MDS values for plotting

tibble(x=mdsvals$x, y=mdsvals$y, sample=rownames(dge.data.TC$samples)) %>%
  inner_join(sample.description) %>%
  ggplot(aes(x=x, y=y, color=trt, shape=sp)) +
  geom_point(size=3) 
```

```{r}
TC.counts.normal <- cpm(dge.data.TC) 

# or log2 transformed:
TC.counts.normal.log <- cpm(dge.data.TC,log = TRUE)
```

```{r}
design.TC <- model.matrix(~sp+trt,data = sample.description)
rownames(design.TC) <- sample.description$sample
design.TC
```

```{r}
design.TC.interaction <- model.matrix(~sp*trt,data = sample.description)
rownames(design.TC) <- sample.description$sample
design.TC
```

```{r}
#First the overall dispersion
dge.data.TC <- estimateGLMCommonDisp(dge.data.TC,design.TC,verbose = TRUE)

#Then a trended dispersion based on count level
dge.data.TC <- estimateGLMTrendedDisp(dge.data.TC,design.TC)

#And lastly we calculate the gene-wise dispersion, using the prior estimates to "squeeze" the dispersion towards the common dispersion.
dge.data.TC <- estimateGLMTagwiseDisp(dge.data.TC,design.TC)

#We can examine this with a plot
plotBCV(dge.data.TC)
```

```{r}
#First the overall dispersion
dge.data.TC.interaction <- estimateGLMCommonDisp(dge.data.TC,design.TC.interaction,verbose = TRUE)

#Then a trended dispersion based on count level
dge.data.TC.interaction <- estimateGLMTrendedDisp(dge.data.TC.interaction,design.TC.interaction)

#And lastly we calculate the gene-wise dispersion, using the prior estimates to "squeeze" the dispersion towards the common dispersion.
dge.data.TC.interaction <- estimateGLMTagwiseDisp(dge.data.TC.interaction,design.TC.interaction)

#We can examine this with a plot
plotBCV(dge.data.TC.interaction)
```

```{r}
#this is the full model with all coefficients
#dge.data is the normalized count data
fit.TC <- glmFit(dge.data.TC, design.TC)

#and for the interaction:
fit.TC.interaction <- glmFit(dge.data.TC.interaction, design.TC.interaction)

#this model looks at genes differentially expressed between two treatments
trt.lrt.TC <- glmLRT(fit.TC,coef = "trtT")

#comparing sp
sp.lrt.TC.TC <- glmLRT(fit.TC,coef = "spTC")
sp.lrt.TC.ST <- glmLRT(fit.TC, coef = "spST")

#interaction
lrt.interaction.TC <- glmLRT(fit.TC.interaction, coef = "spST:trtT")
lrt.interaction.TC2 <- glmLRT(fit.TC.interaction, coef = "spTC:trtT")
```

```{r}
topTags(trt.lrt.TC)
topTags(sp.lrt.TC.TC)
topTags(sp.lrt.TC.ST)
topTags(lrt.interaction.TC)
topTags(lrt.interaction.TC2)
```

```{r}
summary(decideTestsDGE(trt.lrt.TC, p.value = 0.05))
summary(decideTestsDGE(sp.lrt.TC.ST, p.value=0.05))
summary(decideTestsDGE(sp.lrt.TC.TC, p.value=0.05))
summary(decideTestsDGE(lrt.interaction.TC, p.value=0.05))
summary(decideTestsDGE(lrt.interaction.TC2, p.value=0.05))
```

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\--

TC to do:

visualization

annotation

figure out interaction DEGs (same as PM w multiple levels of factors)

lskfjlksdjflksj

\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\-\--
