---
title: "Final DFD Analysis"
author: "Britt"
date: "2026-04-07"
output: html_document
---

```{r}
PM.counts <- read.table("Psme_RNASeq_July2024_Psme.1_0_counts.txt", header = TRUE) 
PM.counts
```

```{r}
PM.counts <- PM.counts[rowSums(PM.counts[,-1] > 10) >= 3,]
PM.counts
```

```{r}
library(tibble)
PM.counts.noid <- as_tibble(PM.counts)
PM.counts.noid
```

```{r}
library(tidyverse)
sample.description <- tibble(sample=colnames(TC.counts.noid))
sample.description <- sample.description %>%
  mutate(
    myc = c("NM", "NM", "NM", "NM", "M", "NM", "M", "M", "M", "NM", "M", "NM", "NM", "M", "NM", "M", "NM", "NM", "NM", "M", "NA"),  
    trt = c("C", "C", "T", "C", "C", "T", "T", "T", "C", "C", "C", "C", "T", "T", "C", "C", "C", "T", "T", "T", "NA"),  
    group = paste(myc, trt, sep = "_")
  )
sample.description <- sample.description %>%
  slice(1:(n() - 1))
sample.description <- sample.description %>%
  mutate(myc=factor(myc, levels = c("NM","M")), 
         trt=factor(trt,levels = c("C","T"))) # setting the levels in this way makes "C and NM" the reference
sample.description
```


```{r}
library(edgeR)

counts.matrix.PM <- PM.counts %>% as.matrix()

#remove unassigned row
counts.matrix.PM <- counts.matrix.PM[, -21]

counts.matrix.PM
```

```{r}
dge.data <- DGEList(counts=counts.matrix.PM, 
                    group=sample.description$group)
dim(dge.data) 
dge.data <- calcNormFactors(dge.data, method = "TMM")
dge.data$samples # look at the normalization factors
```

(make ordination here)

```{r}
# Ensure levels are set correctly
sample.description$myc <- relevel(factor(sample.description$myc), ref = "NM")
sample.description$trt <- relevel(factor(sample.description$trt), ref = "C")
```


```{r}
#sample.description is the original descriptive dataset (describes our samples but doesn't have counts data)

interaction.design <- model.matrix(~myc*trt,data = sample.description)
rownames(interaction.design) <- sample.description$sample
interaction.design
```
```{r}
# Modern, clean workflow for the Interaction model
dge.data <- estimateDisp(dge.data, interaction.design)

# View the dispersion estimates
summary(dge.data$tagwise.dispersion)
```



```{r}
# 1. Fit using the QL model instead of glmFit
fit <- glmQLFit(dge.data, interaction.design)

# 2. Test using the QL F-test instead of glmLRT
# Treatment Effect
trt.qlf <- glmQLFTest(fit, coef = "trtT")

# Mycorrhization Effect (spM)
myc.qlf <- glmQLFTest(fit, coef = "mycM")

# Interaction Effect (The Synergy)
interaction.qlf <- glmQLFTest(fit, coef = "mycM:trtT")
```


```{r}
summary(decideTests(myc.qlf, p.value=0.06))
summary(decideTests(trt.qlf, p.value=0.01))
summary(decideTests(interaction.qlf, p.value=0.5))
```

```{r}
top10.trt <- topTags(trt.qlf)
top10.myc <- topTags(myc.qlf) 
top10.interaction <- topTags(interaction.qlf)

top10.trt
top10.myc
top10.interaction
```
Notes:

interaction: no sig genes, top genes at 0.5 FDR and higher (not 0.25 threshold Laura and I discussed)
myc: sig genes at above 0.05 FDR, set at 0.06 produces sig DEGs
trt: sig genes at 0.01 FDR


```{r}
# Visualizing the dispersion/BCV
plotBCV(dge.data)
```

```{r}
# This will show you if you have an outlier sample
plotMDS(dge.data, col=as.numeric(sample.description$myc))
```

```{r}
mdsvals <- plotMDS(dge.data, plot = FALSE) #using this one for paper

var1 <- round(mdsvals$var.explained[1] * 100, 1)
var2 <- round(mdsvals$var.explained[2] * 100, 1)

tibble(
  x = mdsvals$x,
  y = mdsvals$y,
  sample = rownames(dge.data$samples)
) %>%
  inner_join(sample.description, by = "sample") %>%
  ggplot(aes(x = x, y = y, color = trt, shape = myc)) +
  geom_point(size = 3) +
  labs(
    x = paste0("Dimension 1 (", var1, "%)"),
    y = paste0("Dimension 2 (", var2, "%)"),
    color = "Drought",
    shape = "Mycorrhization"
  ) +
  theme_classic()
```

Notes: PCA does not change with new analysis. Still not looking great.

------------------------------------------------
9/7/26 let's make heatmaps

let's start by getting ALL DEGs from the three treatments

```{r}
all.myc.degs <- as.data.frame(topTags(myc.qlf, n = Inf))
all.trt.degs <- as.data.frame(topTags(trt.qlf, n = Inf))
all.int.degs <- as.data.frame(topTags(interaction.qlf, n = Inf))
```

```{r}
print(all.myc.degs)
print(all.trt.degs)
print(all.int.degs)
```


```{r}
library(edgeR)
library(pheatmap)
```

```{r}
# Filter for significant DEGs (FDR < 0.05) and select top 25 by absolute logFC
sig.trt.degs <- subset(all.trt.degs, FDR < 0.01)
top25.trt.degs <- sig.trt.degs[order(abs(sig.trt.degs$logFC),decreasing = TRUE), ][1:25, ]

sig.myc.degs <- subset(all.myc.degs, FDR < 0.06)
top25.myc.degs <- sig.myc.degs[order(abs(sig.myc.degs$logFC),decreasing = TRUE), ][1:25, ]

top25.int.degs <- all.int.degs[order(abs(all.int.degs$logFC),decreasing = TRUE), ][1:25, ]
```

```{r}
top25.int.degs
top25.myc.degs
top25.trt.degs
```




# -----------------------------------------------------------------------------
# 2. Expression Matrix Preparation & Sample Metadata
# -----------------------------------------------------------------------------
# Get logCPM values for all samples
logcpm <- cpm(dge_object, log = TRUE)

# Subset logCPM matrix to the top 50 DEG gene names
top50_gene_names <- rownames(top50_degs)
heatmap_matrix <- logcpm[top50_gene_names, ]

# Extract sample group annotations directly from DGEList object
annotation_col <- data.frame(
  Group = dge_object$samples$group,
  row.names = colnames(heatmap_matrix)
)

# -----------------------------------------------------------------------------
# 3. Generate Heatmap
# -----------------------------------------------------------------------------
pheatmap(
  heatmap_matrix,
  scale = "row",                      # Standardize row expression (Z-score)
  annotation_col = annotation_col,   # Display group color bars on columns
  show_rownames = TRUE,              # Show top 50 gene IDs
  show_colnames = TRUE,              # Show sample IDs
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Top 50 DEGs by Absolute logFC (FDR < 0.05)"
)
```




