---
title: "trt.sp.annotation"
author: "Britt"
date: "2025-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Annotating the top 29 Treatment (trt), Species (sp), and interaction DEGs

```{r}
int.DEGs <- read.csv("PM.interaction.DEGs.csv")
int.DEGs

sp.DEGs <- read.csv("PM.interaction.DEGs.csv") 
sp.DEGs

trt.DEGs <- read.csv("PM.interaction.DEGs.csv")
trt.DEGs
```


trying the CDS file instead...

```{r}
library(Biostrings)

# Read the CDS FASTA file
cds_sequences <- readDNAStringSet("Psme.1_0.cds.fa.gz")

# Print the sequences to inspect
cds_sequences
```


let's try to extract sequence from specific genes:

```{r}
DEG1 <- cds_sequences["PSME_21469"]
print(DEG1)
```


let's attempt a for loop to automate this:

```{r}

extracted_sequences <- character(nrow(DEGs))

# Loop over each row in the DEGs data frame to extract sequences
for (i in 1:nrow(DEGs)) {
  gene_name <- DEGs[i, 1]  # Extract the gene name from the first column
  
  # Check if the gene name exists in the cds_sequences
  if (gene_name %in% names(cds_sequences)) {
    DEG_sequence <- cds_sequences[gene_name]  # Extract sequence
    DEG_sequence <- as.character(DEG_sequence)  # Convert to character string
  } else {
    DEG_sequence <- NA  # Assign NA if gene is not found
  }
  
  # Store the result in the vector
  extracted_sequences[i] <- DEG_sequence
}

# Add the extracted sequences to the DEGs data frame
DEGs$sequence <- extracted_sequences

# Save the updated DEGs data frame with sequences to a CSV file
write.csv(DEGs, "DEGs_with_sequences.csv", row.names = FALSE)

cat("Sequences have been extracted and saved to 'DEGs_with_sequences.csv'.")

```


That seems to have worked, although I need to double check accuracy later.

Let's remove some columns from the dataframe and then resave the data as a csv.

```{r}
library(dplyr)

# Remove columns by position
DEGs <- DEGs %>% select(-c(3, 4, 5))
DEGs
```


now I have a csv with the 29 DEGs, their sequence, their FDRs and log fold change:

```{r}
write.csv(DEGs, "interaction.DEGs.csv", row.names = FALSE)
```

time to interpro...






